Fractional Types

### 2013-02-04 frac.tex
sec3

In the computer science community, the theory and practice of type
isomorphisms is well-established. Originally, such type isomorphisms were
motivated the pragmatic concern of searching large libraries of functions by
providing one of the many possible isomorphic types for the desired
function. More recently, type isomorphisms have taken a more central role as
\emph{the} fundamental computational mechanism from which more conventional,
i.e., irreversible computation, is derived~\cite{infeffects}. In more detail,
in our own previous work, we started with the notion of type isomorphism as
the ``right'' starting point for a theory of computation. The central
technical contribution was the development of a family of programming
languages, {{langRev}} with various superscripts, in which every computation
is an isomorphism. Other major results are that all such computations
preserve the information-theoretic entropy, that the language is logically
reversible, and that it has a model in symmetric monoidal categories (which
are the backbone of models for linear logic and quantum computation). We have
also invested a significant amount of time in using {{langRev}} as an actual
programming language and established several idioms for programming large
classes of interesting (recursive) programs in {{langRev}}.

@@

We review our language {{langRev}} providing the necessary background and
context for our higher-order extension.\footnote{The presentation in this
  section focuses on the simplest version of {{langRev}}. Other versions
  include the empty type, recursive types, and trace operators but these
  extensions are orthogonal to the higher-order extension emphasized in this
  paper.} The terms of {{Pi}} are not classical values and functions; rather,
the terms are isomorphism witnesses.  In other words, the terms of {{Pi}} are
proofs that certain ``shapes of values'' are isomorphic.  And, in classical
Curry-Howard fashion, our operational semantics shows how these proofs can be
directly interpreted as actions on ordinary values which effect this shape
transformation. Of course, ``shapes of values'' are very familiar already:
they are usually called \emph{types}.  But usually one designs a type system
as a method of classifying terms, with the eventual purpose to show that
certain properties of well-typed terms hold, such as safety.  Our approach is
different: we start from a type system, and then present a term language
which naturally inhabits these types, along with an appropriate operational
semantics.

\paragraph*{Data.}
We view {{Pi}} as having two levels:  it has traditional values, given by
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
and these are classified by ordinary types
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b

@@

\paragraph*{Isomorphisms.} The terms of {{Pi}} are witnesses to type
isomorphisms.  They have (iso) types {{b <-> b}}.  
Specifically, they are witnesses to the following type isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
Each line of the above table introduces a pair of dual constants%
\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness the type
isomorphism in the middle.  These are the base (non-reducible) terms of the
second, principal level of {{Pi}}.

@@

\noindent
To summarize, the syntax of {{Pi}} is given as follows.

\begin{definition}{(Syntax of {{Pi}})}
\label{def:Pi}
We collect our types, values, and isomorphisms, to get the full language
definition.
%subcode{bnf} include main
% value types, b ::= 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= swap+ | assocl+ | assocr+ 
%     &|& unite | uniti | swap* | assocl* | assocr* 
%     &|& dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% top level term, l ::= c v
\end{definition}

The semantics of {{langRev}} is given using two mutually recursive
interpreters: one going forward and one going backwards. The use of {{sym}}
switches control from one evaluator to the other. We will not present the
semantics until the next section in the context of the extended language.
For now, we state without proof that the evaluation of well-typed combinators
always terminates and that {{langRev}} is logically reversible, i.e., that
for all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}}
we have the forward evaluation of {{c v1}} produces {{v2}} iff the backwards
evaluation of {{c v2}} produces {{v1}}.

@@

\begin{definition}[Syntax of {{langRevTF}}]
\label{def:langRevT}
The syntax of types, values, and combinators, is:
%subcode{bnf} include main
% value types, b ::= 1 | b+b | b*b | 1/b 
% values, v ::= () | left v | right v | (v,v) | 1/v 
%
% iso.~types, t ::= b <-> b
% base iso ::= swap+ | assocl+ | assocr+ 
%                   &|& unite | uniti | swap* | assocl* | assocr* 
%                   &|& dist | factor 
%                   &|& eta*_b | eps*_b
% combinators, c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
\end{definition}

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. We think of a value like {{1/left ()}} as a
\emph{first-class constraint} that can only be satisfied if it is matched
with an actual value {{left ()}} or as representing the absence of some
information (i.e., \emph{negative information}), that can only be supplied by
{{left ()}}. The combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission
point which creates --- out of nothing --- an equal amount of negative and
positive information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}}
consumes an equal amount of negative and positive information producing 
no residual information.

@@

>>>>

In the computer science community, the theory and practice of type
isomorphisms is well-established. Originally, such type isomorphisms were
motivated by the pragmatic concern of searching large libraries of functions
by providing one of the many possible isomorphic types for the desired
function. More recently, type isomorphisms have taken a more central role as
\emph{the} fundamental computational mechanism from which more conventional,
i.e., irreversible computation, is derived~\cite{infeffects}. In more detail,
in our own previous work, we started with the notion of type isomorphism as
the ``right'' starting point for a theory of computation. The central
technical contribution was the development of a family of programming
languages, {{langRev}} with various superscripts, in which every computation
is an isomorphism. Other major results are that all such computations
preserve the information-theoretic entropy, that the language is logically
reversible, and that it has a model in symmetric monoidal categories (which
are the backbone of models for linear logic and quantum computation). We have
also invested a significant amount of time in using {{langRev}} as an actual
programming language and established several idioms for programming large
classes of interesting (recursive) programs in {{langRev}}.

@@

We review our language {{langRev}} providing the necessary background and
context for our higher-order extension.\footnote{The presentation in this
  section focuses on the simplest version of {{langRev}}. Other versions
  include the empty type, recursive types, and trace operators but these
  extensions are orthogonal to the higher-order extension emphasized in this
  paper.} The terms of {{langRev}} are not classical values and functions;
rather, the terms are isomorphism witnesses.  In other words, the terms of
{{langRev}} are proofs that certain ``shapes of values'' are isomorphic.
And, in classical Curry-Howard fashion, our operational semantics shows how
these proofs can be directly interpreted as actions on ordinary values which
effect this shape transformation. Of course, ``shapes of values'' are very
familiar already: they are usually called \emph{types}.  But usually one
designs a type system as a method of classifying terms, with the eventual
purpose to show that certain properties of well-typed terms hold, such as
safety.  Our approach is different: we start from a type system, and then
present a term language which naturally inhabits these types, along with an
appropriate operational semantics.

\paragraph*{Data.}
We view {{langRev}} as having two levels:  it has traditional values, given by
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
and these are classified by ordinary types
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b

@@

\paragraph*{Isomorphisms.} The terms of {{langRev}} are witnesses to type
isomorphisms.  They have (iso) types {{b <-> b}}.  
Specifically, they are witnesses to the following type isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
Each line of the above table introduces a pair of dual constants%
\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness the type
isomorphism in the middle.  These are the base (non-reducible) terms of the
second, principal level of {{langRev}}.

@@

\noindent
To summarize, the syntax of {{langRev}} is given as follows.

\begin{definition}{(Syntax of {{langRev}})}
\label{def:Pi}
We collect our types, values, and isomorphisms, to get the full language
definition.
%subcode{bnf} include main
% value types, b ::= 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= swap+ | assocl+ | assocr+ 
%     &|& unite | uniti | swap* | assocl* | assocr* 
%     &|& dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% top level term, l ::= c v
\end{definition}

Mathematically, {{langRev}} can be interpreted in the category of finite sets
and bijections: each type denotes a finite set of the corresponding size and
each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
denoted by {{b1}} and {{b2}}. Operationally, the semantics of {{langRev}} is
given using two mutually recursive interpreters: one going forward and one
going backwards. The use of {{sym}} switches control from one evaluator to
the other. We will not present the operational semantics until the next
section in the context of the extended language. For now, we state without
proof that the evaluation of well-typed combinators always terminates and
that {{langRev}} is logically reversible, i.e., that for all combinators 
{{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have the forward
evaluation of {{c v1}} produces {{v2}} iff the backwards evaluation of 
{{c v2}} produces {{v1}}.

@@

\begin{definition}[Syntax of {{langRevTF}}]
\label{def:langRevT}
We extend Def.~\ref{def:Pi} as follows:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
%
% base iso ::= ... | eta*_b | eps*_b
\end{definition}

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. We think of a value like {{1/left ()}} as a
\emph{first-class constraint} that can only be satisfied if it is matched
with an actual value {{left ()}} or as representing the absence of some
information (i.e., \emph{negative information}), that can only be supplied by
{{left ()}}. The combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission
point which creates --- out of nothing --- an equal amount of negative and
positive information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}}
matches up an equal amount of negative and positive information producing 
no residual information.

%%%%%%%%%%%%%%%%%%%%%5
\subsection{Relational Semantics}

Even though the categorical semantics of {{langRevTF}} is not the main focus
of this paper, we find it helpful to briefly outline an interpretation of
{{langRevTF}} in \emph{dagger compact closed categories}. These categories
were introduced by Abramsky and
Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and then by
Selinger~\cite{Selinger:2007:DCC:1229185.1229207}. The simplest example of
such categories is the category of finite sets and relations on which we
focus.

We recall that a \emph{symmetric monoidal category} is a category together
with a bifunctor $\otimes$, a distinguished object $I$, and natural
isomorphisms $\alpha_{A,B,C} : (A \otimes B) \otimes C \rightarrow A \otimes
(B \otimes C)$, $\lambda_A : A \rightarrow I \otimes A$, and $\sigma_{A,B} :
A \otimes B \rightarrow B \otimes A$ subject to standard coherence
conditions~\cite{nla.cat-vn1051288}. Following common practice
(e.g.,~\cite{Selinger:2007:DCC:1229185.1229207}), we write $\rho_A =
\sigma_{I,A} \circ \lambda_A : A \rightarrow A \otimes I$.

A \emph{compact closed category} is a symmetric monoidal category where each
object $A$ is assigned a dual object $A^*$, together with a unit map $\eta_A
: I \rightarrow A^* \otimes A$ and a counit map $\epsilon_A : A \otimes A^*
\rightarrow I$, such that:
\[\begin{array}{rcl}
\lambda_A^{-1} \circ (\epsilon_A \otimes A) \circ \alpha_{A,A^*,A}^{-1} \circ (A \otimes \eta_A) \circ \rho_A &=& \mathit{id}_A \\
\rho_{A^*}^{-1} \circ (A^* \otimes \epsilon_A) \circ \alpha_{A^*,A,A^*} \circ (\eta_A \otimes A^*) \circ \lambda_A &=& \mathit{id}_{A^*}
\end{array}\]

the category of finite sets and relations.

@@

model using vector spaces; category of finite sets and relations; homotopy
equivalence???
